Mathematical appendix to "Spatio-temporal estimates of HIV risk group proportions for adolescent

### 2022-12-17 src/docs_paper/appendix.Rmd
A few small appendix edits

Let $i \in \{1, \ldots, n\}$ denote subnational units (Table B.5) which partition the 13 studied AGYW priority countries $c[i] \in \{1, \ldots, 13\}$.
We considered the years 1999-2018 denoted as $t \in \{1, \ldots, T\}$ and age groups $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$.
Let the four risk groups be $k \in \{1, 2, 3, 4\}$ and denote being in either the third or fourth risk group by $k = 3^{+}$.

First, we used a multinomial logistic regression model to infer the proportion of AGYW in the risk groups $k \in \{1, 2, 3^{+}\}$.
This model can be specified using a multinomial likelihood
\begin{equation}
    \y_{ita} = (y_{ita1}, \ldots, y_{ita3^{+}})^\top \sim \text{Multinomial}(m_{ita}; \, p_{ita1}, \ldots, p_{ita3^{+}}),
\end{equation}
where the number of women in risk group $k$ is $y_{itak}$, the sample size is $m_{ita} = \sum_{k = 1}^{3^{+}} y_{itak}$, and $p_{itak} > 0$ is the probability of membership of the $k$th risk group with $\sum_{k = 1}^{3^{+}} p_{itak} = 1$.
Taking $k = 1$ to be the baseline category, linear predictors may be specified for the additive log ratios $\log(p_{itak} / p_{ita1})$.
Rather than taking this approach directly, to facilitate inference in the `R-INLA`, we used an equivalent Poisson log-linear model via the multinomial-Poisson transformation [@baker1994multinomial].
This transformation, and further details about this model, are presented in Section 1.2.

Next, we fit a logistic regression model to estimate the proportion of those in the $k = 3^{+}$ risk group that were in the $k = 3$ and $k = 4$ risk groups respectively.
This model was of the form
\begin{align}
    y_{ia4} &\sim \text{Binomial} \left( y_{ia3} + y_{ia4}, q_{ia} \right), \label{eq:logistic-regression} \\
    q_{ia} &= \text{logit}^{-1} \left( \eta_{ia} \right), 
\end{align}
where $q_{ia} = p_{ia4} / (p_{ia3} + p_{ia4}) = p_{ia4} / p_{ia{3^+}}$ and the linear predictor $\eta_{ia}$ is chosen suitably, described in more detail in Section 1.3.
Taking this two-step approach allowed us to include all surveys in the multinomial regression model, but only those surveys with a specific transactional sex question in Equation \ref{eq:logistic-regression}.
As all such surveys occurred in 2013-2018, in the logistic regression model $q_{ia}$ was assumed to be constant over time.

To facilitate uncertainty quantification, we took 1000 posterior samples (indexed by $s$) from each of the multinomial $\{p_{itak}^s\}$ and logistic regression $\{q_{ia}^s\}$ models.
Samples $p_{ita3}^s$ and $p_{ita4}^s$ were then generated by splitting samples from the $k = 3^{+}$ category of the multinomial model via $p_{ita3}^s = (1 - q_{ia}^s)p_{ita{3^+}^s}$ and $p_{ita4}^s = q_{ia}^s p_{ita{3^+}}^s$.

As transactional sex does not directly correspond to sex work [@wamoyi2016transactional], we adjusted our samples so that at a national level, the population size estimates match FSW population size estimates at a national level by age.
Our methodology for producing these estimates is described in Section \ref{fsw-population-size-estimation}.
In making this adjustment, we assumed that subnational variation in the FSW proportions corresponds to that of the transactional sex proportions.

We calculated the mean, median, 2.5\% and 97.5\% quantiles for the risk group probabilities empirically using the adjusted samples.
To produce aggregated estimates, such as the age category 15-24 or any national estimates, we weighted the adjusted samples by population sizes $N_{ita} = \sum_k N_{itak}$ obtained from the Naomi model [@eaton2021naomi].

@@

\begin{table}[htb]
\small
\centering
\begin{tabularx}{\textwidth}{lllllll}
\midrule
Model ID & Category ($\beta_k$) & Country ($\zeta_{ck}$) & Age ($\alpha_{ack}$) & Spatial ($\phi_{ik}$) & Temporal ($\gamma_{tk}$) & Spatio-temporal ($\delta{itk}$) \\
\midrule
M1 & IID & IID & IID & IID & IID & Not included \\
M2 & IID & IID & IID & Besag & IID & Not included \\
M3 & IID & IID & IID & IID & AR1 & Not included \\
M4 & IID & IID & IID & Besag & AR1 & Not included \\
\midrule
\end{tabularx}
\caption{The multinomial regression models that we considered. Observation random effects $\theta_{ita}$ are omitted from this table.}
\label{tab:multinomial-models}
\small
\end{table}

@@

We considered four models (Table \ref{tab:multinomial-models}) for $\eta_{ita}$ of the form
$$
\eta_{ita} = \theta_{ita} + \beta_k + \zeta_{c[i]k} + \alpha_{ac[i]k} + \phi_{ik} + \gamma_{tk}.
$$
<!-- + \delta_{itk} -->
As described above, we included observation random effects $\theta_{ita} \sim \mathcal{N}(0, 1000^2)$ in all models we considered.
To capture country-specific proportion estimates for each category, we included category random effects $\beta_k \sim \mathcal{N}(0, \tau_\beta^{-1})$ and country-category random effects $\zeta_{ck} \sim \mathcal{N}(0, \tau_\zeta^{-1})$.
Heterogeneity in risk group proportions by age was allowed by including age-country-category random effects $\alpha_{ack} \sim \mathcal{N}(0, \tau_\alpha^{-1})$.
We considered two specifications, independent and identically distributed (IID) and Besag [@besag1991bayesian], for the space-category $\phi_{ik}$ random effects (Section \ref{sec:spatial-random-effects}) and two specifications, IID and first order autoregressive (AR1), for the year-category $\gamma_{tk}$ random effects (Section \ref{sec:temporal-random-effects}).
<!-- We also considered separable space-year-category interactions $\delta_{itk}$ with corresponding Kronecker product structure (Section \ref{sec:spatio-temporal-random-effects}). -->
All random effect precision parameters $\tau \in \{\tau_\beta, \tau_\zeta, \tau_\alpha, \tau_\phi, \tau_\gamma\}$ were given independent penalised complexity (PC) priors [@simpson2017penalising] with base model $\sigma = 0$ given by $p(\tau) = 0.5 \nu \tau^{-3/2} \exp \left( - \nu \tau^{-1/2} \right)$ where $\nu = - \ln(0.01) / 2.5$ such that  $\mathbb{P}(\sigma > 2.5) = 0.01$.
<!-- \tau_\delta -->

@@

The surveys we included use a complex design, in which each individual has an unequal probability of being included in the sample.
The DHS [@measure2012sampling], for example, often employs a two-stage cluster design, first taking an urban rural stratified sample of ennumeration areas, before selecting households from each ennumeration area using systematic sampling.

@@

We performed model selection on the basis of the conditional predictive ordinate (CPO) criterion [@pettit1990conditional], selecting model M2.
This model included Besag spatial random effects and IID temporal random effects.
The deviance information criterion (DIC) [@spiegelhalter2002bayesian] and widely applicable information criterion (WAIC) [@watanabe2013widely] are included for reference (Table \ref{tab:multinomial-model-performance} and Figure \ref{fig:multinomial-model-performance}).

@@

Next, we used the assumed $\text{Gamma}(\alpha = 10.4, \beta = 0.36)$ FSW age distribution in South Africa from the Thembisa model [@johnson2020thembisa] to calculate the implied ratio between the number of FSW and the sexually debuted population in each age group.
We assumed that those ratios in South Africa were applicable to every country, and thereby calculated the number of FSW by age group in all 13 countries (Figure \ref{fig:age-disagg-fsw-line}).

@@

We calculated HIV prevalence $\rho_{iak}$ and number of people living with HIV (PLHIV) $H_{iak}$ stratified according to district, age group and risk group by disaggregating Naomi estimates by risk group.

>>>>

Let $i \in \{1, \ldots, n\}$ denote subnational units (Table B.5) which partition the 13 studied AGYW priority countries $c[i] \in \{1, \ldots, 13\}$.
We considered the years 1999-2018 denoted as $t \in \{1, \ldots, T\}$, and age groups $a \in \{\text{15--19}, \text{20--24}, \text{25--29}\}$.
Let the four risk groups be $k \in \{1, 2, 3, 4\}$ and denote being in either the third or fourth risk group by $k = 3^{+}$.

First, we used a multinomial logistic regression model to infer the proportion of AGYW in the risk groups $k \in \{1, 2, 3^{+}\}$.
One way to specify this model is using a multinomial likelihood
\begin{equation}
    \y_{ita} = (y_{ita1}, \ldots, y_{ita3^{+}})^\top \sim \text{Multinomial}(m_{ita}; \, p_{ita1}, \ldots, p_{ita3^{+}}),
\end{equation}
where the number of women in risk group $k$ is $y_{itak}$, the sample size is $m_{ita} = \sum_{k = 1}^{3^{+}} y_{itak}$, $p_{itak} > 0$ is the probability of membership of the $k$th risk group with $\sum_{k = 1}^{3^{+}} p_{itak} = 1$, and taking $k = 1$ to be the baseline category, linear predictors may be specified for the additive log ratios $\log(p_{itak} / p_{ita1})$.
To facilitate inference in the `R-INLA`, we used an equivalent Poisson log-linear model via the multinomial-Poisson transformation [@baker1994multinomial].
This transformation, and further details about this model, are presented in Section 1.2.

Next, we fit a logistic regression model to estimate the proportion of those in the $k = 3^{+}$ risk group that were in the $k = 3$ and $k = 4$ risk groups respectively.
This model was of the form
\begin{align}
    y_{ia4} &\sim \text{Binomial} \left( y_{ia3} + y_{ia4}, q_{ia} \right), \label{eq:logistic-regression} \\
    q_{ia} &= \text{logit}^{-1} \left( \eta_{ia} \right), 
\end{align}
where $q_{ia} = p_{ia4} / (p_{ia3} + p_{ia4}) = p_{ia4} / p_{ia{3^+}}$ and the linear predictor $\eta_{ia}$ is chosen suitably, described in more detail in Section 1.3.
Taking this two-step approach allowed us to include all surveys in the multinomial regression model, but only those surveys with a specific transactional sex question in Equation \ref{eq:logistic-regression}.
As all such surveys occurred in 2013-2018, in the logistic regression model we assumed $q_{ia}$ to be constant over time.

To facilitate uncertainty quantification, we took 1000 posterior samples, indexed by $s$, from each of the multinomial $\{p_{itak}^s\}$ and logistic regression $\{q_{ia}^s\}$ models.
Samples $p_{ita3}^s$ and $p_{ita4}^s$ were then generated by $p_{ita3}^s = (1 - q_{ia}^s)p_{ita{3^+}^s}$ and $p_{ita4}^s = q_{ia}^s p_{ita{3^+}}^s$.

As transactional sex does not directly correspond to sex work [@wamoyi2016transactional], we adjusted our samples so that at a national level, the population size estimates match FSW population size estimates by age.
Our methodology for producing these estimates is described in Section \ref{fsw-population-size-estimation}.
In making this adjustment, we assumed that subnational variation in the FSW proportions corresponds to that of the transactional sex proportions.

We calculated the mean, median, and 2.5\% and 97.5\% quantiles for the risk group probabilities empirically using the adjusted samples.
To produce aggregated estimates, such as the age category 15-24 or any national estimates, we weighted the adjusted samples by population sizes $N_{ita} = \sum_k N_{itak}$ obtained from the Naomi model [@eaton2021naomi].

@@

\begin{table}[htb]
\small
\centering
\begin{tabularx}{\textwidth}{lllllll}
\midrule
Model ID & Category ($\beta_k$) & Country ($\zeta_{ck}$) & Age ($\alpha_{ack}$) & Spatial ($\phi_{ik}$) & Temporal ($\gamma_{tk}$) & Spatio-temporal ($\delta{itk}$) \\
\midrule
M1 & IID & IID & IID & IID & IID & Not included \\
M2 & IID & IID & IID & Besag & IID & Not included \\
M3 & IID & IID & IID & IID & AR1 & Not included \\
M4 & IID & IID & IID & Besag & AR1 & Not included \\
\midrule
\end{tabularx}
\caption{The multinomial regression models that we considered. Observation random effects $\theta_{ita}$, included in all models, are omitted from this table.}
\label{tab:multinomial-models}
\small
\end{table}

@@

We considered four models (Table \ref{tab:multinomial-models}) for $\eta_{ita}$ of the form
$$
\eta_{ita} = \theta_{ita} + \beta_k + \zeta_{c[i]k} + \alpha_{ac[i]k} + \phi_{ik} + \gamma_{tk}.
$$
<!-- + \delta_{itk} -->
Observation random effects $\theta_{ita} \sim \mathcal{N}(0, 1000^2)$ were included in all models we considered.
To capture country-specific proportion estimates for each category, we included category random effects $\beta_k \sim \mathcal{N}(0, \tau_\beta^{-1})$ and country-category random effects $\zeta_{ck} \sim \mathcal{N}(0, \tau_\zeta^{-1})$.
Heterogeneity in risk group proportions by age was allowed by including age-country-category random effects $\alpha_{ack} \sim \mathcal{N}(0, \tau_\alpha^{-1})$.
We considered two specifications, independent and identically distributed (IID) and Besag [@besag1991bayesian], for the space-category $\phi_{ik}$ random effects (Section \ref{sec:spatial-random-effects}) and two specifications, IID and first order autoregressive (AR1), for the year-category $\gamma_{tk}$ random effects (Section \ref{sec:temporal-random-effects}).
<!-- We also considered separable space-year-category interactions $\delta_{itk}$ with corresponding Kronecker product structure (Section \ref{sec:spatio-temporal-random-effects}). -->
All random effect precision parameters $\tau \in \{\tau_\beta, \tau_\zeta, \tau_\alpha, \tau_\phi, \tau_\gamma\}$ were given independent penalised complexity (PC) priors [@simpson2017penalising] with base model $\sigma = 0$ given by $p(\tau) = 0.5 \nu \tau^{-3/2} \exp \left( - \nu \tau^{-1/2} \right)$ where $\nu = - \ln(0.01) / 2.5$ such that  $\mathbb{P}(\sigma > 2.5) = 0.01$.
<!-- \tau_\delta -->

@@

We included surveys which use a complex design, in which each individual has an unequal probability of being included in the sample.
For example the DHS often employs a two-stage cluster design, first taking an urban rural stratified sample of ennumeration areas, before selecting households from each ennumeration area using systematic sampling [@measure2012sampling].

@@

We performed model selection on the basis of the conditional predictive ordinate (CPO) criterion [@pettit1990conditional], selecting model M2.
This model included Besag spatial random effects and IID temporal random effects.
For reference, we also computed the deviance information criterion (DIC) [@spiegelhalter2002bayesian] and widely applicable information criterion (WAIC) [@watanabe2013widely] (Table \ref{tab:multinomial-model-performance} and Figure \ref{fig:multinomial-model-performance}).

@@

Next, we used the assumed $\text{Gamma}(\alpha = 10.4, \beta = 0.36)$ FSW age distribution in South Africa from the Thembisa model [@johnson2020thembisa] to calculate the implied ratio between the number of FSW and the sexually debuted population in each age group.
We assumed these ratios in South Africa were applicable to every country to calculate the number of FSW by age group in all 13 countries (Figure \ref{fig:age-disagg-fsw-line}).

@@

We calculated HIV prevalence $\rho_{iak}$ and the number of people living with HIV (PLHIV) $H_{iak}$ stratified according to district, age group and risk group by disaggregating Naomi estimates by risk group.
